\psection{Partitioning}

\textbf{Goal:} Partition \emph{cells} into \emph{blocks} $\implies$ Graph
partitioning.\\
Given a graph $G(V,E)$, with $|V|$ nodes and $|E|$ edges.\\
Each node has area $s(v)$ and each edge has cost/weight $w(e)$.\\
Divide graph $G$ into $k$ disjoint subgraphs. (NP-hard)\\
Optimization goals:
\begin{itemize}
  \item Minimum number of connections between partitions
  \item Design constraints (size, external connections)
  \item Balanced partition
\end{itemize}

\describe{Kernighan-Lin Algorithm} \\
Given: Graph with $2n$ nodes, same weight. \\
Goal: Partition into $A$ and $B$; min. cost, $|A|=|B|=n$.\\
*Cost $D(v)$ of moving a node $v$
$$D(v)=|E_c(v)|-|E_{nc}(v)|$$
$$ E_c(v)/E_{nc}(v) = \text{edges of $v$ that are (not) cut} $$
*Gain of swapping a pair of nodes $a$ and $b$
$$ \Delta g = D(a) + D(b) - 2 \cdot c(a,b) $$
$$ D(a),D(b) = \text{costs of nodes $a,b$} $$
$$ c(a,b) = \text{connection weight between $a$ and $b$} $$
*Maximum positive gain $G_m$ of a pass
$$ G_m = \sum_{i=1}^{m} \Delta g_i $$
Algorithm: $O(n^3)$ per pass\\
\begin{algorithm}[H]
\Repeat( \tcp*[h]{a single pass}){no improvement ($G_m < 0$)}{
  \Repeat{each node swapped over}{
    calculate $D(v)$ for all nodes\;
    find max. $\Delta g_i$ and swap\;
  }
  select max. positive gain $G_m = \sum_{i=1}^{m} \Delta g_i$\;
  perform $m$ moves\;
}
\end{algorithm}
Extensions:
\begin{itemize}
  \item unequal partition size
  \item unequal node weights
  \item $k$-way partitioning (recursive)
\end{itemize}

\describe{Fiduccia-Mattheyses Algorithm}
\begin{itemize}
  \item swapping becomes moving nodes
  \item slight imbalance allowed
  \item edges become nets (w/ 2+ pins)
  \item Node = cell; subgraph = block
\end{itemize}
Given: Hypergraph $G(V,H)$, weighted hyperedges\\
Goal: Partition int $k$, min. weight cost\\
*Gain $\Delta g(c)$ for cell $c$
$$ \Delta g(c) = FS(c) - TE(c) $$
$$ FS(c) = \text{moving force, i.e. cut nets only connected to $c$} $$
$$ TE(c) = \text{retention force, i.e. uncet nets connected to $c$} $$
*Maximum positive gain $G_m$ of a pass
$$ G_m = \sum_{i=1}^{m} \Delta g_i $$
*Ratio factor, i.e. relative balance between two partitions
$$ r = \frac{area(A)}{area(A)+area(B)} $$
*Balance criterion  
$$ area(V) \in [ r \cdot area(V) \pm area_{max}(V) ] $$
*Base cell: Cell $c$ with max. cell gain $\Delta g(c)$ among all free cells
and does not violate balance criterion\\
Algorithm: $O(n)$ per pass \quad(sorted node list)\\
\begin{algorithm}[H]
  \Repeat( \tcp*[h]{a single pass}){no improvement}{
    compute balance criterion\;
    \Repeat{all cells are fixed}{
      find max. $\Delta g_i$ and fix cell $c_i$\;
    }
    select max. positive gain $G_m = \sum_{i=1}^{m} \Delta g_i$\;
    perform $m$ moves\;
  }
\end{algorithm}

\describe{Multilevel Partitioning}\\
Clustering: ($\implies$ Simulated Annealing)
\begin{itemize}
  \item group tighlty-connected nodes into clusters
  \item partition clusters
  \item refine partition by splitting clusters
\end{itemize}
